In this exercise, you will practice basic `dplyr` (the relevant package within `tidyverse`) operations on a dataset of gene expression levels across human tissues (from the GTEx Project). The dataset (`dt`) has one row per gene, with columns:

-   `Name`: the Ensembl gene ID\
-   `Description`: the gene symbol
-   One column per tissue containing the median expression levels (in TPM units)

------------------------------------------------------------------------

### Start by loading the data into R:

Note the arguments, which specify that the first two lines of the file will be skipped, that the file contains a header line that should be used to form column names, and that the columns are tab (`\t`) separated.

```{r}
library(tidyverse)
dt <- read_delim("/Users/cmdb/Data/GTEx/GTEx_Analysis_2017-06-05_v8_RNASeQCv1.1.9_gene_median_tpm.gct.gz", skip = 2, col_names = TRUE, delim = "\t")

print(dt)
```

### Q1. How many genes are in the dataset, and how many tissues are represented?

```{r}
#answers

dim(dt)

#genes in dataset:56,200 (# of rows in the tibble) 
#tissues that are represented: 54 tissues


```

------------------------------------------------------------------------

### Q2. How many genes are expressed above a given threshold in each tissue?

A common threshold is 1 TPM to define whether a gene is "expressed."\
Compute, for each of the three tissues, `Liver`, `Testis`, and `Whole Blood`, how many genes are expressed above this threshold.\
Then identify which tissue has the **most** and which has the **fewest** expressed genes. See this paper for one possible explanation: Xia et al. 2021: [PMC7891839](https://pmc.ncbi.nlm.nih.gov/articles/PMC7891839/).

Do the results change if you set the threshold higher, say at 10 TPM?

```{r}

dt %>%
  filter(Liver > 1) %>%
  select(Name, Liver) %>%
  dim()

dt %>%
  filter(Testis > 1) %>%
  select(Name, Testis) %>%
  dim()

dt %>%
  filter(`Whole Blood` > 1) %>%
  select(Name, `Whole Blood`) %>%
  dim()


dt %>%
  filter(Liver > 0) %>%
  select(Name, Liver) %>%
  dim()

dt %>%
  filter(Testis > 0) %>%
  select(Name, Testis) %>%
  dim()

dt %>%
  filter(`Whole Blood` > 0) %>%
  select(Name, `Whole Blood`) %>%
  dim()

dt %>%
  filter(Liver > 10) %>%
  select(Name, Liver) %>%
  dim()

dt %>%
  filter(Testis > 10) %>%
  select(Name, Testis) %>%
  dim()

dt %>%
  filter(`Whole Blood` > 10) %>%
  select(Name, `Whole Blood`) %>%
  dim()

# Interpretation here as a comment
#NUMBER OF GENES EXPRESSED
#Liver: 13499
#Testis: 25064
#Whole Blood: 11616

#Tissue with most expressed genes: Testis
#Tissue with least expressed genes: Whole Blood 

# >10 for expression 
#Liver: 5132
#Testis: 11471
#Whole Blood: 4482
# the tissue with the most expressed genes and least expressed genes remained the same, but the number of genes decreased. ________ 


  
```

------------------------------------------------------------------------

### Q3. What are the 20 most highly expressed genes in the liver?

What do you notice about many of the genes in this list? Can you propose a hypothesis to explain your observation?

```{r}

high_exp_genes <- dt %>%
  select(Name, Liver) %>%
  arrange(desc(Liver)) %>%
  head(n = 20)

# Interpretation here as a comment
# 20 most highly expressed genes in liver:
#  1 ENSG00000198886.2  58775.
#  2 ENSG00000198899.2  51821.
#  3 ENSG00000198712.1  49564.
#  4 ENSG00000198938.2  44571.
#  5 ENSG00000198804.2  40519.
#  6 ENSG00000198727.2  31642.
#  7 ENSG00000228253.1  29932.
#  8 ENSG00000212907.2  26086.
#  9 ENSG00000163631.16 25201.
# 10 ENSG00000198763.3  22892.
# 11 ENSG00000198888.2  20800.
# 12 ENSG00000198840.2  19264.
# 13 ENSG00000210082.2  18806.
# 14 ENSG00000257017.8  12259.
# 15 ENSG00000173432.10  8842 
# 16 ENSG00000229314.5   8806.
# 17 ENSG00000087086.14  7718.
# 18 ENSG00000110245.11  7600.
# 19 ENSG00000171564.11  7352.
# 20 ENSG00000198786.2   6072.



```

------------------------------------------------------------------------

### Q4. Compare the mean expression levels of mitochondrial versus nuclear-encoded genes in the Liver.

Genes encoded by the mitochondrial genome have the gene symbol prefix `MT-`. The function `str_starts()` from the `stringr` package in `tidyverse` tests whether a string (or a vector of strings) start with a given pattern, returning TRUE/FALSE (or a boolean vector) as output.

Use `str_starts()` to identify which genes are mitochondrially encoded and store this information in a new Boolean column named `is_mitochondrial`. Next, compute the mean and median expression of mitochondrial genes in the liver, as well as the mean and median expression of nuclear-encoded genes in the liver and compare them. Do mitochondrial genes exhibit higher or lower mean and median expression than nuclear genes? What does the difference between the mean and median tell you about the distribution of gene expression.

```{r}


mit_genes <- dt %>%
  select(Description, Liver) %>%
  mutate(is_mitochondrial = str_starts(Description, "MT-")) %>%
  filter(is_mitochondrial == TRUE)


nuc_genes <- dt %>%
  select(Description, Liver) %>%
  mutate(is_mitochondrial = str_starts(Description, "MT-")) %>%
  filter(is_mitochondrial == FALSE)


mean_mit <- mean(mit_genes$Liver)
median_mit <- median(mit_genes$Liver)
mean_nuc <- mean(nuc_genes$Liver)
median_nuc <- median(nuc_genes$Liver)

print(mean_mit)
print(median_mit)
print(mean_nuc)
print(median_nuc)


  
# Interpretation here as a comment
#The mean expression of mitochondrial genes in the liver is about a thousand times more than that of the nuclear-encoded genes, whereas the medians are the same. This means that there is a much wider range of values of mitochondrial genes, and there are some outliers which very higher expression pulling the mean to be larger, because the median is still lower. 



```

------------------------------------------------------------------------

### Q5. Are liver’s most highly expressed genes also highly expressed in another tissue?

Take the 20 most highly exprssed genes in **Liver** and check how they are expressed in the **Heart - Left Ventricle**. Identify:\
1. one gene that is highly expressed in both tissues\
2. one gene that is highly expressed in Liver but very low in Heart

Look up basic information about these genes on the internet. Do their expression patterns make sense given their known functions?

```{r}

# Code here 
liver_genes_in_heart_ventricle <- dt %>%
  select(Description, Liver, `Heart - Left Ventricle`) %>%
  arrange(desc(Liver)) %>%
  head(n = 20)

# Interpretation here as a comment
#One gene that is highly expressed in both the liver and left ventricle of the heart is . And one gene that is expressed highly in the liver but lowly in the heart is . 

```

------------------------------------------------------------------------

### Q6. Comparing gene expression between tissues

Another way to compare tissues is to ask how similar are their expression patterns overall, across all genes?

R has a built-in function called `cor.test()` that measures the correlation between two vectors of numbers. Remember that you can extract the contents of a given column of a tibble as a vector using the syntax `tibble_name$column_name`. If a column name contains spaces, enclose the column name in backticks such as `` tibble_name$`column name` ``.

Use `cor.test()` with the argument `method = "kendall"` to compute: - the Spearman correlation between `Liver` and `Heart - Left Ventricle` - the Spearman correlation between `Liver` and `Brian - Cortex`

Which pair of tissues is more strongly correlated? Does this result make sense biologically?

```{r}
# Liver vs Heart answer

liver_vector <- dt$Liver 
heart_vector <- dt$`Heart - Left Ventricle`
liver_heart_cor <- cor.test(liver_vector, heart_vector, method = "kendall")
print(liver_heart_cor)


# Liver vs Brain answer
liver_vector <- dt$Liver 
brain_vector <- dt$`Brain - Anterior cingulate cortex (BA24)`
liver_brain_cor <- cor.test(liver_vector, brain_vector, method = "kendall")
print(liver_brain_cor)

# interpretation
#Liver Heart cor: tau = 0.8053218 

#Liver Brain cor: tau = 0.733215 



```

------------------------------------------------------------------------

### Optional: further exploration

What other patterns are present in the data? Use tidyverse functions to explore and summarize the data in different ways and see what you can learn.

```{r}
# answer here
```
